
{\centering\par
\Large\scshape\chapTitleFont\thetitle
\par}
\vspace*{2\baselineskip}

{%
\setlength{\parskip}{1.5em}
\setlength{\parindent}{0pt}

The material in this book has been edited from talks given at Chithurst Forest Monastery in January 1984, with the exception of three sections. These are `\textit{Effort and Relaxation},' `\textit{Kindness},' `\textit{The Hindrances and Their Cessation},' which were taken from talks given at Wat Pah Nanachat International Forest Monastery, Ubon, N.E. Thailand in December 1982.

The photographs are of slabs from the ruins of the ancient \textit{st\=upa} at Amaravati in Andhra Pradesh, India; they are reproduced by kind permission of the Trustees of the British Museum, London.

Page \pageref{image-stupa}: the \textit{st\=upa}, a monumental reliquary, contains the relics of a saint. As the object of pilgrimages, it symbolizes the universal quest for spiritual truth.

Page \pageref{image-feet}: the iconographical footprints of the Buddha. They represent the path that a teacher has taken, to be followed by his disciples.

Page \pageref{image-lotus-scene}: the lotus of wisdom in a scene of everyday human activity.

\clearpage
\thispagestyle{empty}

{\scshape \theauthor} was ordained as a Therav\=adin \textit{bhikkhu} in Thailand, 1966. He spent ten years in N.E. Thailand training under the Venerable Ajahn Chah, a master of the Forest Tradition. In 1976 he was invited to Britain by the English Sangha Trust, and has subsequently established two major monasteries -- Cittaviveka and Amaravati, as well as smaller branch monasteries. He was abbot of Amaravati Buddhist Monastery just north of London from its consecration in 1984 until his retirement in 2010.

{\scshape Ajahn Sucitto} is the abbot of Cittaviveka, Chithurst Buddhist Monastery since 1992. He contributed the Introduction to this booklet in 1986. He was born in London, and was ordained in Thailand in March 1976. He moved to Britain in 1978 and took up training under Ajahn Sumedho at the Hampstead Buddhist Vihara. In 1979 he was one of the small group of monks, led by Ajahn Sumedho, who established Cittaviveka, Chithurst Buddhist Monastery, in West Sussex.


}

